Dynamic Witnesses for Static Type Errors

### 2015-11-20 paper/pldi16-submission/evaluation.tex

We answer these questions on two sets of known-bad programs, \ie
programs that were rejected by the \ocaml compiler because of a type
error.
%
The first dataset comes from the Spring 2014 undergraduate Programming
Languages course at our institution.
%
We recorded each interaction with the \ocaml top-level system over the
course of the first three assignments (IRB \# hidden for blind review), % \#140608),
from which we extracted 4,407 distinct, ill-typed \ocaml programs.
%
The second dataset -- widely used in the literature -- comes from a
similar course at the University of
Washington~\cite{lerner_seminal:_2006}, from which we extracted 284
ill-typed programs.

>>>>

We answer these questions on two sets of known-bad programs, \ie
programs that were rejected by the \ocaml compiler because of a type
error.
%
The first dataset comes from the Spring 2014 undergraduate Programming
Languages course at our institution.
%
We recorded each interaction with the \ocaml top-level system over the
course of the first three assignments (IRB \# hidden for blind review), % \#140608),
from which we extracted \ucsdsize\ distinct, ill-typed \ocaml programs.
%
The second dataset -- widely used in the literature -- comes from a
similar course at the University of
Washington~\cite{lerner_seminal:_2006}, from which we extracted 284
ill-typed programs.

### 2015-11-20 paper/pldi16-submission/introduction.tex

\paragraph{3. Evaluating Witnesses}
%
Our third contribution is an extensive evaluation of our approach
on two different sets of ill-typed programs obtained by instrumenting
compilers used in beginner's classes~(\S~\ref{sec:eval}).
%
The first is the \uwbench\ data set~\cite{lerner_searching_2007}
comprising \uwsize\ ill-typed programs.
%
The second is a new \ucsdbench\ data set, comprising \ucsdsize\
ill-typed programs.
%
We show that for both data sets, our technique is able to generate
witnesses for nearly 90\% of the programs, in under a second the
vast majority of cases.
%
Furthermore, we show that a simple interactive strategy yields
compact counterexample traces with fewer than 5 steps for 60\%
of the programs, and under 10 steps for 90\% of the programs.

In short, our results show that at least for beginners, ill-typed
programs \emph{do} go wrong, in the vast majority of cases, which,
in turn, opens the door to a completely different, dynamic way
to explain, understand, and appreciate the benefits of static typing.

>>>>

\paragraph{3. Evaluating Witnesses}
%
Our third contribution is an extensive evaluation of our approach
on two different sets of ill-typed programs obtained by instrumenting
compilers used in beginner's classes~(\S~\ref{sec:evaluation}).
%
The first is the \uwbench\ data set~\cite{lerner_searching_2007}
comprising \uwsize\ ill-typed programs.
%
The second is a new \ucsdbench\ data set, comprising \ucsdsize\
ill-typed programs.
%
We show that for both data sets, our technique is able to generate
witnesses for nearly 90\% of the programs, in under a second the
vast majority of cases.
%
Furthermore, we show that a simple interactive strategy yields
compact counterexample traces with fewer than 5 steps for 60\%
of the programs, and under 10 steps for 90\% of the programs.

\smallskip
In short, our results show that at least for beginners, ill-typed
programs \emph{do} go wrong, in the vast majority of cases, which,
in turn, opens the door to a completely different, dynamic way
to explain, understand, and appreciate the benefits of static typing.

### 2015-11-20 paper/pldi16-submission/overview.tex

We start with an overview of our approach to
explaining (static) type errors using \emph{witnesses}
that (dynamically) show how the program goes wrong.
%
We illustrate why generating suitable inputs
to functions is tricky in the absence of type
information.
%
Then we describe our solution to the problem
and highlight the similarity to static type
inference,
%
Finally, we demonstrate our visualization of the witnesses we produce.

@@

\paragraph{Problem: Which inputs are bad?}
%
One approach is to randomly generate input values and
use them to execute the program until we find one that
causes the program to go wrong.
%
Unfortunately, this approach quickly runs aground.
Consider the @fac@ function from Figure~\ref{fig:factorial},
reproduced here.

\begin{code}
  let rec fac n =
    if n <= 0 then
      true
    else
      n * fac (n-1)
\end{code}
\ES{having two copies of \texttt{fac} seems silly, but the back-reference across a page boundary is no good either...}

What \emph{types} of inputs should we test @fac@ with?
%
Values of type @int@ are fair game, but values of type, say,
@string@ or @int list@ will cause the program to go wrong
in an \emph{irrelevant} manner.
%
Concretely, we want to avoid testing @fac@ with any type other
than @int@ because any other type would cause @fac@ to get stuck
immediately in the @n <= 0@ test.

>>>>

We start with an overview of our approach to
explaining (static) type errors using \emph{witnesses}
that (dynamically) show how the program goes wrong.
%
We illustrate why generating suitable inputs
to functions is tricky in the absence of type
information.
%
Then we describe our solution to the problem
and highlight the similarity to static type
inference,
%
Finally, we demonstrate our visualization of
the synthesized witnesses.

@@

\paragraph{Problem: Which inputs are bad?}
%
One approach is to randomly generate input values and
use them to execute the program until we find one that
causes the program to go wrong.
%
Unfortunately, this approach quickly runs aground.
Recall the erroneous @fac@ function from Figure~\ref{fig:factorial}.
%~\S~\ref{sec:introduction}:
%
% \begin{code}
  % let rec fac n =
    % if n <= 0 then
      % true
    % else
      % n * fac (n-1)
% \end{code}
% \ES{having two copies of \texttt{fac} seems silly, but the back-reference across a page boundary is no good either...}
%
What \emph{types} of inputs should we test @fac@ with?
%
Values of type @int@ are fair game, but values of type, say,
@string@ or @int list@ will cause the program to go wrong
in an \emph{irrelevant} manner.
%
Concretely, we want to avoid testing @fac@ with any type other
than @int@ because any other type would cause @fac@ to get stuck
immediately in the @n <= 0@ test.
